cardinalR: Generating interesting high-dimensional data structures

### 2025-09-23 paper-cardinalR.Rmd
Combine first two paragraphs of introduction
Did all the comments for introduction

Generating synthetic datasets with clearly defined geometric properties is useful for evaluating and benchmarking algorithms in various fields, such as machine learning, data mining, and computational biology. Researchers often need to generate data with specific dimensions, noise characteristics, and complex underlying structures to test the performance and robustness of their methods.

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The `geozoo` package (@barret2016) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. The package is `snedata` (@james2025), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, `splatter` (@luke2017) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression. In contrast, `mlbench` (@friedrich2024) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges.

While these packages are valuable, their scope is often limited to specific applications or low-dimensional structures.

To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By using clearly defined geometric shapes and controllable characteristics such as number of dimensions, sample size; `cardinalR` allows researchers to generate transparent and interpretable synthetic datasets for evaluating algorithm performance in high-dimensional settings.

By providing these aspects, `cardinalR` provides researchers with a method to generate more explainable and challenging high-dimensional clustering synthetic datasets focused to the specific needs of evaluating algorithms.

>>>>

Generating synthetic datasets with clearly defined geometric properties is useful for evaluating and benchmarking algorithms in various fields, such as machine learning, data mining, and computational biology. Researchers often need to generate data with specific dimensions, noise characteristics, and complex underlying structures to test the performance and robustness of their methods. There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The `geozoo` package (@barret2016) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. The package is `snedata` (@james2025), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, `splatter` (@luke2017) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression. In contrast, `mlbench` (@friedrich2024) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges.

While these packages are valuable, their scope is often limited to specific applications or low-dimensional structures. To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By using clearly defined geometric shapes and controllable characteristics such as number of dimensions, sample size; `cardinalR` allows researchers to generate transparent and interpretable synthetic datasets useful for evaluating the performance of nonlinear dimensionality reduction (NLDR) methods, clustering algorithms, and visualization techniques. Moreover, these datasets can serve as benchmark examples for exploring how different algorithmic choices affect the identification or representation of cluster and manifold structures in high-dimensional spaces.

### 2025-09-23 paper-cardinalR.tex
Combine first two paragraphs of introduction
Did all the comments for introduction

Generating synthetic datasets with clearly defined geometric properties is useful for evaluating and benchmarking algorithms in various fields, such as machine learning, data mining, and computational biology. Researchers often need to generate data with specific dimensions, noise characteristics, and complex underlying structures to test the performance and robustness of their methods.

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The \texttt{geozoo} package (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. The package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{splatter} (\citet{luke2017}) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression. In contrast, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges.

While these packages are valuable, their scope is often limited to specific applications or low-dimensional structures.

To address this gap, this paper introduces the \texttt{cardinalR} R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. \texttt{cardinalR} offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By using clearly defined geometric shapes and controllable characteristics such as number of dimensions, sample size; \texttt{cardinalR} allows researchers to generate transparent and interpretable synthetic datasets for evaluating algorithm performance in high-dimensional settings.

By providing these aspects, \texttt{cardinalR} provides researchers with a method to generate more explainable and challenging high-dimensional clustering synthetic datasets focused to the specific needs of evaluating algorithms.

>>>>

Generating synthetic datasets with clearly defined geometric properties is useful for evaluating and benchmarking algorithms in various fields, such as machine learning, data mining, and computational biology. Researchers often need to generate data with specific dimensions, noise characteristics, and complex underlying structures to test the performance and robustness of their methods. There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The \texttt{geozoo} package (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. The package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{splatter} (\citet{luke2017}) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression. In contrast, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges.

While these packages are valuable, their scope is often limited to specific applications or low-dimensional structures. To address this gap, this paper introduces the \texttt{cardinalR} R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. \texttt{cardinalR} offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By using clearly defined geometric shapes and controllable characteristics such as number of dimensions, sample size; \texttt{cardinalR} allows researchers to generate transparent and interpretable synthetic datasets useful for evaluating the performance of nonlinear dimensionality reduction (NLDR) methods, clustering algorithms, and visualization techniques. Moreover, these datasets can serve as benchmark examples for exploring how different algorithmic choices affect the identification or representation of cluster and manifold structures in high-dimensional spaces.
